Mining HIV Trends in Social Media Data

### 2016-04-19 draft/draft.tex
finally locked in all results, can now focus entirely on writing a draft, cool

Pre-Exposure Prophylaxis (PrEP) is a recently developed therapy for the prevention of Human Immunodeficiency Virus (HIV). The treatment is available as a pill named Truvada and is taken once a day for as much as 99\% protection from infection when exposed to HIV. While medically effective, PrEP suffers from high drug costs, HIV-related social stigma, and under-informed health providers and patients. This paper contributes a data mining and machine learning strategy that monitors Twitter social media data. Our results show that we can identify trends over time, and positive and negative PrEP related tweets. These results can be used by public health professionals to inform and protect at risk individuals from HIV.

@@

We excluded non-English tweets, and also tweets that did not originate in contiguous United States time zones. A subset of the tweets we collected had geolocation data available. We found that tweets were largely concentrated on US and Canadian metro areas (Figure 1).

@@

% TODO work this section Tuesday

The first analyses that we performed sought to identify certain keywords, hashtags, tweets and users, that were discussing HIV and PrEP related contexts. Word2Vec and the related method, Paragraph2Vec are unsupervised machine learning methods that have performed well at embedding natural language in a semantic vector space. In our analysis, Word2Vec allows us to determine semantically similar words to a query word, while Paragraph2Vec allows us to determine similar tweets, users and hashtags to a query hashtag.

% want to do a tSNE plot, but only on things that are relatively dissimilar. For example do HIV, truvada, STDs, testing related, awareness related, ... and throw away or merge labels

% table 1 - w2v PrEP
\begin{table}
\centering
\caption{Cosine Similarity to 'PrEP'}
\begin{tabular}{|l|c|} \hline
Related Word & Cosine Similarity to 'PrEP'\\ \hline
\hline\end{tabular}
\end{table}

% table 2 - d2v PrEP
\begin{table}
\centering
\caption{Cosine Similarity to '\#truvada'}
\begin{tabular}{|l|c|} \hline
Related Hashtag/Tweet & Cosine Similarity to '\#truvada'\\ \hline
\hline\end{tabular}
\end{table}

% show PCA for each table above (remember to mention percent varience)

@@

Next we sought to identify some temporal trends in PrEP related trends. We used Dynamic Topic Modelling (DTM) to identify how certain topics change over time. In DTM, documents are grouped in several corpuses that represent different points in time. Latent Dirichlet Analysis (LDA) is then performed on each corpus, to extract a topic distribution. The posterior distribution from time $t_n$ is used as the prior for time $t_{n+1}$. We specified 10 topics and used each week's worth of tweets over a 30 week period as our time points.

% might want to explain at a high level how DTM works here

We identified two topics that showed relevant trends. In topic 5 we see that the keyword 'prep' remains relatively constant, while 'prevention' and 'truvada' decline over time. 'hivaids' increases and then decreases in magnitude over the course of the time series (Figure 3). These dynamics may indicate short term events in hivaids and PrEP-related conversation on twitter over the time period studied.

% figure 3
\begin{figure}
\centering
\includegraphics[height=2.5in, width=3.5in]{DTMfig1}
\caption{DTM topic 5, word prevalence over time. Date is YYYY-WW.}
\end{figure}

We found that topic 4 captured several keywords related to World AIDS Day (Figure 4). We can see that all of these terms peak in the 47th week of 2015 and then decline into 2016. This correlates well with the actual date of World AIDS Day, December 1st. Furthermore,while December 1st is World AIDS day, the whole month of December is AIDS Awareness month. We are clearly seeing the word 'awareness' peak later and last longer that the word 'WorldAIDSDay' implying that it is correlated to the whole month of December. While we aren't specifically interested in World AIDS Day, or AIDS awareness month, to inform our understanding of PrEP discussion, this observation validates our ability to identify temporal events using DTM.

% figure 4
\begin{figure}
\centering
\includegraphics[height=2.5in, width=3.5in]{DTMfig2}
\caption{DTM topic 5, word prevalence over time. Date is YYYY-WW.}
\end{figure}

These results demonstrate our ability to extract relevant HIV and PrEP related information from twitter that accurately captures time-dependant fluctuations. Public health professionals should be able to monitor these temporal trends to determine the relative interest in PrEP related social media discourse. This could lead to a system for early detection of HIV outbreak detection, or other serious developments.

@@

% figure 5
\begin{figure}
\centering
\includegraphics[height=2.5in, width=3.5in]{user_timeline_clustermap}
\caption{Clustermap of users vs. topics for the top 300 users related to PrEP.}
\end{figure}

% table 3
\begin{table*}
\centering
\caption{Topic modeling on PrEP-related users\' timelines.}
\begin{tabular}{|l|c|c|c|c|c|c|c|c|c|c|} \hline
Topic Number & Word 0 & Word 1 & Word 2 & Word 3 & Word 4 & Word 5 & Word 6 & Word 7 & Word 8 & Word 9\\ \hline
0 & hivaids & condoms & bitcoin & stigma & cdcstd & kanyewest & nyhaad & entrepreneur & realestate & blacklivesmatter\\ \hline
1 & ang & ng & tcot & ako & sa & ko & pjnet & realdonaldtrump & gopdebate & cruzcrew\\ \hline
2 & fertility & infertility & meme & mxico & rp & deactivate & rail & roc & offensive & feedly\\ \hline
3 & colombia & pozlivingbc & fe & repostapp & sprite & bccfe & thewalkingdead & npinchat & interior & wellness\\ \hline
4 & businessinsider & stacks & noh & teaparty & sk & holi & alzheimers & caregiver & asa & guru\\ \hline
5 & armenia & wordpress & georgia & topping & lgbtqia & yah & faction & checkout & narendramodi & examination\\ \hline
6 & knitting & lista & naral & gust & worldbank & bigdata & haiku & vdeo & mhealth & hah\\ \hline
7 & nigga & niggas & bernie & lmao & nhsengland & bitches & bae & uniteblue & il & gop\\ \hline
8 & socialmedia & seo & socialmediamarketing & der & und & entrepreneur & ist & nowplaying & foxbusiness & enews\\ \hline
9 & collage & mileycyrus & undetectable & ddlovato & casa & cnet & qotd & vacations & nashville & heros \\ \hline

@@

% do a top positive tweets table (also with tweet ID and log prob of classification)

% do a top negative tweets table (also with tweet ID and log prob of classification)

>>>>

Pre-Exposure Prophylaxis (PrEP) is a recently developed therapy for the prevention of Human Immunodeficiency Virus (HIV). The treatment is available as a pill named Truvada and is taken once a day for as much as 99\% protection from infection when exposed to HIV. While medically effective, PrEP suffers from high drug costs, HIV-related social stigma, and under-informed health providers and patients. This paper contributes a data mining and machine learning strategy that monitors Twitter social media data. Our results show that we can identify PrEP and HIV discussion trends over time, and positive and negative PrEP related tweets. These results can be used by public health professionals to inform and protect at-risk individuals from HIV.

@@

We excluded tweets that did not originate in contiguous United States time zones. A subset of the tweets we collected had geolocation data available. We found that tweets were largely concentrated on US and Canadian metro areas (Figure 1).

@@

The first analysis that we performed sought to identify certain keywords, hashtags, tweets and users, that were discussed in HIV and PrEP related contexts. Word2Vec and the related method, Paragraph2Vec are unsupervised machine learning methods that have performed well at embedding natural language in a semantic vector space. In our analysis, Word2Vec allows us to determine semantically similar words to a query word, while Paragraph2Vec allows us to determine similar tweets, users and hashtags to a query hashtag.

% table 1 - w2v PrEP
\begin{table}
\centering
\caption{Cosine Similarity to word-vector 'PrEP'}
\begin{tabular}{|l|c|} \hline
Related Word & Cosine Similarity to 'PrEP'\\ \hline
truvada & 0.796666\\ \hline
DoingIt & 0.738141\\ \hline
WorldAIDSDay & 0.720910\\ \hline
NancyReagan & 0.717667\\ \hline
NBHAAD & 0.705061\\ \hline
ART & 0.704300\\ \hline
HLM2016AIDS & 0.698698\\ \hline
ICASA2015 & 0.693117\\ \hline
HIV & 0.692860\\ \hline
OneConversation & 0.688427\\ \hline
\hline\end{tabular}
\end{table}

% table 2 - d2v PrEP
\begin{table}
\centering
\caption{Cosine Similarity to doc-vector '\#PrEP'}
\begin{tabular}{|l|c|} \hline
Related Hashtag/Tweet & Cosine Similarity to '\#PrEP'\\ \hline
\#lgbtmedia16 & 0.739128\\ \hline
\#hiv & 	0.727602 \\ \hline
\#whereisprep & 0.707165 \\ \hline
\#truvada & 0.696113 \\ \hline
\#hivprevention & 0.636068 \\ \hline
tweet-702179860983189504 & 0.630055\\ \hline
user-711275699529764864 & 0.629254\\ \hline
tweet-708519265540907010 & 0.628778 \\ \hline
tweet-712032637024653313 & 0.628646 \\ \hline
\#harrogatehour & 0.628547 \\ \hline
\hline\end{tabular}
\end{table}

% example tweets that show up related to Prep that are informative, and misinformative:
% tweet-702179860983189504 "#StoneColdVideoTODAY if You see this 13 symptoms. Do HIV Test Immediately. Must Read"
% tweet-708519265540907010 "Checkout why PrEP is hurting the cause amp # JoinTheConversation # LGBTQIA"

% figure 3
\begin{figure*}
\centering
\includegraphics[height=4in, width=6in]{tSNE_word}
\caption{tSNE plot of relevant word-vectors}
\end{figure*}

@@

Next we sought to identify some temporal trends in PrEP related trends. We used Dynamic Topic Modelling (DTM) to identify how certain topics change over time. In DTM, documents are grouped in several corpuses that represent successive points in time. Latent Dirichlet Analysis (LDA) is then performed on each corpus, to extract a topic distribution. The posterior topic distribution from time $t_n$ is used as the prior for time $t_{n+1}$. This lets the model determine a topic model for each time point that is dependent both on the set of tweets from that time point, and on the time point's document distribution. We specified 10 topics and used each week's worth of tweets, over a 30 week period from the 47th week of 2015 to the 14th week of 2016, as our time points.

We identified two topics that showed relevant trends. In topic 5 we see that the keyword 'PrEP' increases over time, while 'prevention', 'drug' and 'risk' remain constant (Figure 3). The term 'pill', also from topic 5, declines overtime. These dynamics may indicate that PrEP discussion is becoming more prevalent. Other HIV prevention related words such as 'pill', 'prevention' and 'drug' which serve as a form of negative controls. They are related to PrEP semantically, and are thus in the same topic. Yet the fact that they are not increasing, shows us that the increase in PrEP discussion is PrEP-specific.

However, it is hard to tell from these data though, whether the increased level of PrEP discussion is leading to increased levels of informed patients, medical providers, and adherence. It is possible that stigma, and misinformation is leading to greater levels of PrEP discussion on twitter. We will get more specific, granular understanding of the PrEP discourse in our sentiment analysis section below.

% figure 3
\begin{figure}
\centering
\includegraphics[height=2.5in, width=3.5in]{DTMfig1}
\caption{DTM topic 5 (PrEP related topic) word prevalence over time. Date is YYYY-WW.}
\end{figure}

We found at least one other DTM topic that showed interesting behaviour. We found that topic 4 captured several keywords related to World AIDS Day (Figure 4). We can see that 'WorldAIDSDay', and 'can' peak in the 47th week of 2015 and then decline into 2016. This correlates well with the actual date of World AIDS Day, December 1st. Furthermore, while December 1st is World AIDS day, the whole month of December is AIDS Awareness Month. We can clearly see the words 'raise' and 'awareness' peak later and last longer that the word 'WorldAIDSDay' implying that it is correlated to the whole month of December. While our PrEP investigation isn't specifically interested in World AIDS Day, or AIDS awareness month, this observation validates our ability to accurately identify temporal events using DTM.

% figure 4
\begin{figure}
\centering
\includegraphics[height=2.5in, width=3.5in]{DTMfig2}
\caption{DTM topic 4 (WorldAIDSDay related topic) word prevalence over time. Date is YYYY-WW.}
\end{figure}

Together, the DTM results demonstrate our ability to extract relevant HIV and PrEP related information from twitter that accurately captures time-dependant fluctuations. Public health professionals should be able to monitor these temporal trends to determine the relative interest in PrEP, and other HIV related keywords as they are discussed over time. This method could even lead to a system for early detection of HIV outbreaks, or other serious developments.

@@

% figure 5
\begin{figure*}
\centering
\includegraphics[height=2.5in, width=3.5in]{user_timeline_clustermap}
\caption{Clustermap of users vs. topics for the top 300 users related to PrEP.}
\end{figure*}

% table 3
\begin{table*}
\centering
\caption{Topic modeling on PrEP-related users' timelines.}
\begin{tabular}{|l|c|c|c|c|c|c|c|c|c|c|} \hline
Topic Number & Word 0 & Word 1 & Word 2 & Word 3 & Word 4 & Word 5\\ \hline
0 & wordpress & dentists & rb & aquarius & faction & reconnect\\ \hline
1 & empirefox & crooks & csr & payroll & ch & allin\\ \hline
2 & vdeo & pra & gust & lgbtqia & rail & undetectable\\ \hline
3 & hivaids & cdcstd & stigma & blacklivesmatter & realestate & getcovered\\ \hline
4 & nigga & niggas & realdonaldtrump & lmao & cruz & gop\\ \hline
5 & fertility & infertility & armenia & collage & nig & deactivate\\ \hline
6 & packaging & buenas & amigo & bu & birthcontrol & reservations \\ \hline
7 & mga & ka & socialmedia & lang & entrepreneur & naman\\ \hline
8 & infosec & bigdata & multimedia & stl & fx & malware\\ \hline
9 & nhsengland & sosa & tw & knitting & worldbank & thtorguk\\ \hline
\hline\end{tabular}
\end{table*}

@@

% do a top positive tweets table (also with tweet ID and log prob of classification)
% table 4
\begin{table*}
\centering
\caption{Positive Sentiment Tweets.}
\begin{tabular}{|p{2cm}|p{12cm}|} \hline
Category & Text\\ \hline
General & "RT TOPublicHealth The Works provides testing for HIV anonymous \& rapid test available . Call 416-392-0520 for more info"\\ \hline
General & "RT FCAA ejaforg announced 5.4 million in grants to support orgs addressing \#HIV in new \& innovative ways!"\\ \hline
General & "RT HillaryClinton A note on the fight against HIV and AIDS and the people who really started the conversation."\\ \hline

PrEP specific & "He won't use condoms because intimacy means more than his health. but he's discovered PrEP. thank goodness."\\ \hline
PrEP specific & "PrEP Queensland Aids Council, \#HIV Foundation, Queensland."\\ \hline
PrEP specific & "RT JDatTheBody At the core of our programs is belief that young ppl can succeed in take PrEP for HIV prevention. \#NHPC2015"\\ \hline

Truvada specific & "RT CDC\_HIVAIDS Expanding testing, treatment, \& \#PrEP could prevent up to 185k new \#HIV infections"\\ \hline
Truvada specific & "Another reason 4 \#Ireland \& \#UK 2 immediately approve \#truvada \& \#PrEP 2 stop \#HIV infections . arleavitt AodhanORiordain MerchantsQuayIR"\\ \hline
Truvada specific & "RT EvanJPeterson For \#worldAIDSday my early \#PrEP arcticle in strangerslog, art by leviathanleague \#hiv \#truvada \#truvadawhore"\\ \hline

% do a top negative tweets table (also with tweet ID and log prob of classification)
% table 5
\begin{table*}
\centering
\caption{Negative Sentiment Tweets.}
\begin{tabular}{|p{2cm}|p{12cm}|} \hline
Category & Text\\ \hline
General & "Also, how fucking vile of Hillary to say. Reagan did fucking NOTHING during the AIDS epidemic until it was too late. What a stupid old hag."\\ \hline
General & "I wonder why he beat her ass when she was tryna leave like she wasn't gone be running back when she found out she had HIV \& nobody want her"\\ \hline
General & "Aaannd. Hillary Clinton breathes a sigh of relief that Twitter has left its outrage of her AIDS comments behind to tend to Drumpf debacle."\\ \hline

PrEP specific & "RT gaston\_croupier \#Truvada patent's not expired yet but it is sold online as a generic drug? There's something rotten in internet \#PrEP h"\\ \hline
PrEP specific & "Equality\_MI Syph \& Hep C have gone up 550 in Gay Men bc many feel tht bc they're on PrEP, they don't need condoms. HIV isn't the only STI."\\ \hline
PrEP specific & "Xaviom8 in interviews he says he was adherent. strain was highly resistant, and Truvada wouldn't have blocked it anyways. PrEP didn't fail."\\ \hline

Truvada specific & "not surprised at all that someone got HIV on truvada. people get pregnant on birth control. tomato-condoms are still important-tomahto"\\ \hline
Truvada specific & "Now reading that truvada does not protect against certain strains of the HIV virus. Yet people want to take that risk.."\\ \hline
Truvada specific & "I think I have conjunctivitis unless truvada cured it overnight cuz im not feeling as horrible today as last night"\\ \hline

% there is viable information in the LDA topic modeling on positive and negative tweets
